Evolution, functional diversity, regime shifts, hysteresis, and nonlinearity; ecosystem responses to environmental change.

### 2022-01-17 experiments/2_supplement/html_supplement.Rmd
many changes!

The replication method shows the same effects of functional diversity as those revealed by the replication method:

@@

```{r}
rm(list=ls())
library(tidyverse)
library(microxanox)
library(patchwork)
library(here)
source(here("r/various_useful_functions.r"))
colfunc_CB <- colorRampPalette(c("#024F17", "#B5FFC9"))
colfunc_SB <- colorRampPalette(c("#7D1402", "#FCBEB3"))
colfunc_PB <- colorRampPalette(c("#6E0172", "#F9AEFC"))
zero <- 0
unity <- 1
ss_9s <- readRDS(here("experiments/0_ss_finding/temporal_method/data/9_strain_SS_data.RDS"))
sort(unique(ss_9s$CB_var_gmax_s))
sort(unique(ss_9s$SB_var_gmax_s))
ss_result <- ss_9s %>%
  filter(abs(CB_var_gmax_s - 0.028254848) < 0.0001,
         abs(SB_var_gmax_s - 0.084764545) < 0.0001)
parameter <- ss_result$parameter[[1]]
parameter$strain_parameter <- ss_result$pars[[1]]

wait_time <- 1e6
#p$log10a_series <- seq(-2.5, -2, 0.025)
parameter$log10a_series <- seq(-8, 0, length = 300)
parameter$sim_duration <- wait_time * length(parameter$log10a_series)
parameter$sim_sample_interval <- wait_time
total_initial_abundances <- 10^5
parameter <- set_temporal_ssfind_initial_state(parameter, total_initial_abundances, 
                total_initial_abundances, total_initial_abundances)
parameter
check_results <- run_temporal_ssfind_method(parameter)

check_results_1e6 <- readRDS(here("experiments/2_supplement/data/check1.RDS"))
junk <- list()
junk$ssfind_result[[1]] <- check_results_1e6
plot_ss_result3(junk, 1)

@@

# Bibliography

>>>>

The replication method shows the same effects of functional diversity as those revealed by the temporal method:

@@

The results below show the stable states as a function of oxygen diffusivity for a system with intermediate level of diversity in all functional groups. There appears to be no effect of using a longer simulation time. This shows that the duration of the simulations is sufficient to reach stable state, or very close to it.

Below is Figure 3 from the main manuscript, in which the duration at each oxygen diffusivity value is 1e6:

```{r echo = FALSE, fig.height=10}

single_1e6 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/single_1e6_result.RDS"))
# sort(unique(ss_9s$CB_var_gmax_s))
# sort(unique(ss_9s$SB_var_gmax_s))
# ss_result <- ss_9s %>%
#   filter(abs(CB_var_gmax_s - 0.028254848) < 0.0001,
#          abs(SB_var_gmax_s - 0.084764545) < 0.0001,
#          abs(PB_var_gmax_s - 0.084764545) < 0.0001)

temp <- single_1e6 %>%
  mutate(a = 10^a_O) %>%
  # arrange(a) %>%
  # mutate(direction = rep(c("up", "down"), nrow(ss_result)/2)) %>%
  #mutate(direction = ifelse(initial_N_CB == 1, "up", "down")) %>%
  # select(-initial_N_CB, -a_O) %>%
  gather(species, quantity, 2:(ncol(.) - 2)) %>%
  mutate(
    var_type = ifelse(grepl("B_", species), "Organism", "Substrate"),
    functional_group = case_when(
      str_detect(species, "CB_") ~ "CB",
      str_detect(species, "SB_") ~ "SB",
      str_detect(species, "PB_") ~ "PB"
    ),
    log10_quantity = log10(quantity + 1)
  )

num_strains <- temp %>%
  group_by(functional_group) %>%
  summarise(num = length(unique(species))) %>%
  na.omit()

num_CB_strains <- num_strains$num[num_strains$functional_group == "CB"]
num_SB_strains <- num_strains$num[num_strains$functional_group == "SB"]
num_PB_strains <- num_strains$num[num_strains$functional_group == "PB"]

line_width <- 1

p1 <- temp %>%
  dplyr::filter(functional_group == "CB")  %>%
  mutate(species = factor(species, levels = unique(species))) %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  scale_colour_manual(values = colfunc_CB(num_CB_strains)) +
  guides(colour = guide_legend(ncol = 3)) +
  theme(legend.position="none") +
  ggtitle("(b) Cyanobacteria")
#p1# p1

p2 <- temp %>%
  dplyr::filter(functional_group == "SB")  %>%
  mutate(species = factor(species, levels = unique(species))) %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  scale_colour_manual(values = colfunc_SB(num_SB_strains)) +
  guides(colour = guide_legend(ncol = 3)) +
  theme(legend.position="none") +
  ggtitle("(c) Sulfate reducing bacteria")

p3 <- temp %>%
  dplyr::filter(functional_group == "PB")  %>%
  mutate(species = factor(species, levels = unique(species))) %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  scale_colour_manual(values = colfunc_PB(num_PB_strains)) +
  guides(colour = guide_legend(ncol = 3)) +
  theme(legend.position="none") +
  ggtitle("(d) Phototrophic sulfur bacteria")

p4 <- temp %>%
  dplyr::filter(var_type == "Substrate", species == "O") %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  theme(legend.position="none") +
  ggtitle("(e) Oxygen concentration")

p5 <- temp %>%
  dplyr::filter(var_type == "Substrate", species == "SR") %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width, col = 6) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  theme(legend.position="none") +
  ggtitle("(f) Sulfide concentration")

patchwork <- p1 / p2 / p3 / p4 / p5
patchwork
```

For comparison, here are the results with twice the duration at each oxygen diffusivity:

```{r echo = FALSE, fig.height=10}

single_2e6 <- readRDS(here("experiments/0_ss_finding/temporal_method/data/single_1e6_result.RDS"))
# sort(unique(ss_9s$CB_var_gmax_s))
# sort(unique(ss_9s$SB_var_gmax_s))
# ss_result <- ss_9s %>%
#   filter(abs(CB_var_gmax_s - 0.028254848) < 0.0001,
#          abs(SB_var_gmax_s - 0.084764545) < 0.0001,
#          abs(PB_var_gmax_s - 0.084764545) < 0.0001)

temp <- single_1e6 %>%
  mutate(a = 10^a_O) %>%
  # arrange(a) %>%
  # mutate(direction = rep(c("up", "down"), nrow(ss_result)/2)) %>%
  #mutate(direction = ifelse(initial_N_CB == 1, "up", "down")) %>%
  # select(-initial_N_CB, -a_O) %>%
  gather(species, quantity, 2:(ncol(.) - 2)) %>%
  mutate(
    var_type = ifelse(grepl("B_", species), "Organism", "Substrate"),
    functional_group = case_when(
      str_detect(species, "CB_") ~ "CB",
      str_detect(species, "SB_") ~ "SB",
      str_detect(species, "PB_") ~ "PB"
    ),
    log10_quantity = log10(quantity + 1)
  )

num_strains <- temp %>%
  group_by(functional_group) %>%
  summarise(num = length(unique(species))) %>%
  na.omit()

num_CB_strains <- num_strains$num[num_strains$functional_group == "CB"]
num_SB_strains <- num_strains$num[num_strains$functional_group == "SB"]
num_PB_strains <- num_strains$num[num_strains$functional_group == "PB"]

line_width <- 1

p1 <- temp %>%
  dplyr::filter(functional_group == "CB")  %>%
  mutate(species = factor(species, levels = unique(species))) %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  scale_colour_manual(values = colfunc_CB(num_CB_strains)) +
  guides(colour = guide_legend(ncol = 3)) +
  theme(legend.position="none") +
  ggtitle("(b) Cyanobacteria")
#p1# p1

p2 <- temp %>%
  dplyr::filter(functional_group == "SB")  %>%
  mutate(species = factor(species, levels = unique(species))) %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  scale_colour_manual(values = colfunc_SB(num_SB_strains)) +
  guides(colour = guide_legend(ncol = 3)) +
  theme(legend.position="none") +
  ggtitle("(c) Sulfate reducing bacteria")

p3 <- temp %>%
  dplyr::filter(functional_group == "PB")  %>%
  mutate(species = factor(species, levels = unique(species))) %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  scale_colour_manual(values = colfunc_PB(num_PB_strains)) +
  guides(colour = guide_legend(ncol = 3)) +
  theme(legend.position="none") +
  ggtitle("(d) Phototrophic sulfur bacteria")

p4 <- temp %>%
  dplyr::filter(var_type == "Substrate", species == "O") %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  theme(legend.position="none") +
  ggtitle("(e) Oxygen concentration")

p5 <- temp %>%
  dplyr::filter(var_type == "Substrate", species == "SR") %>%
  ggplot(aes(x = log10(a), y = log10_quantity, col = species)) +
  geom_path(lwd = line_width, col = 6) +
  ##ylab("log10(quantity [cells])") +
  ylab("Log(Quantity)") +
  xlab("Oxygen diffusivity") +
  theme(legend.position="none") +
  ggtitle("(f) Sulfide concentration")

patchwork <- p1 / p2 / p3 / p4 / p5
patchwork
```

@@

### 2022-01-17 manuscript (obsolete)/article.tex
many changes!

This report concerns how biodiversity influences the strength and nature
of ecosystem responses to environmental change (figure 1; panel(a) three
ovals, environment oval with arrow to ecosystem property oval, with
biodiversity oval pointing the arrow between environment and ecosystem
property; panel (b) shows linear response with different gradients,
large being with low diversity; panel (c) shows a linear and a
non-linear response starting and ending at the same place; panel (d)
shows a hysteretic response, with greater or less range of hysteresis).
Strong responses of ecosystems to environmental change imply high
\emph{sensitivity} and lack of resistance to environmental change, and
higher temporal variability in the face of varying environmental
conditions. Greater \emph{non-linearity} in ecosystem responses
increases the potential for surprises, such as when an ecosystem that
was resistant to an environmental change becomes unexpectedly sensitive
to further change. The presence and extent of \emph{hysteresis} in
ecosystem responses to environmental driver gradients concerns how
readily are ecosystem responses reversed when environmental change is
reversed.

@@

\begin{itemize}
\item
  In general, greater functional response diversity may decrease the
  likelihood of a catastrophic ecosystem response (Figure 2a in Dakos et
  al. (2019)). Here, we interpret this prediction to be that greater
  functional response diversity decreases the non-linearity of ecosystem
  response to environmental driver change, including the extreme
  non-linearity that can occur when an ecosystem ``flips'' from one
  state to another one.
\item
  Greater functional trait diversity can make a tipping point occur only
  with greater change in an environmental driver (Figure 2a in Dakos et
  al. (2019)). I.e. functional trait diversity can increase
  resilience--the extent of environmental change required to push the
  ecosystem into another state (Holling (1973)). This could occur if
  greater functional diversity equates to the presence of organisms with
  greater tolerance to the environmental driver Dakos et al. (2019).
\item
  On the other hand, the opposite could occur: greater functional
  diversity could make a tipping point occur with less change in an
  environmental driver (i.e.~may cause lower resilience) (Figure 2c in
  Dakos et al. (2019)). This could occur if there is selection for
  greater tolerance that causes, incidentally, selection for lower
  growth rate, perhaps due to a tradeoff between the two traits.
\item
  Finally, the amount of functional diversity may affect the ``recovery
  trajectory,'' i.e.~the relationship between ecosystem state and the
  environmental driver when the driver change is reversed Dakos et al.
  (2019). Put another way, functional diversity may increase or decrease
  the amount of hysteresis, and may even result in a recovery with lower
  non-linearity / no tipping-point (Figure 3 in Dakos et al. (2019)).
\end{itemize}

These explanations/predictions were based on the verbal/graphical
argumentation Dakos et al. (2019). There were no mathematical models of
ecosystem responses to environmental change behind the
explanations/predictions. Ceulemans, Wojcik, and Gaedke (2021), in their
article \emph{Functional diversity alters the effects of a pulse
perturbation on the dynamics of tritrophic food webs}, did employ a such
a mathematical model {[}to be developed{]}.

@@

\begin{itemize}
\tightlist
\item
  To provide an answer to the question of how trait variation affects
  the non-linearity, resilience and hysteresis of ecosystem response to
  an environmental change.
\end{itemize}

@@

\begin{itemize}
\item
  Mathematical modelling of a microbrial ecosystem. Trait diversity
  manipulation (can be considered selection among strains, or selection
  among species). (Do not include mutation or recomination).
\item
  Variation in amount of among strain variation, no variation in number
  of strains, no variation in strength or shape of the of any tradeoff.
\item
  Measurement of the effect on how the steady state(s) of the system
  varies with in different environmental conditions, and on the
  dynamical response of system to gradual environmental change.
\end{itemize}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/owenpetchey/Desktop/microxanox/diversity_envresp1/manuscript/ecosystem_network}

@@

Bush et al (2017) contains an accessible and complete description of the
model of the ecosystem, including rate equations, parameter values, and
initial conditions. These features are also represented and documented
in the \texttt{microxanox} R package (\textbf{cite the package}).
Therefore, here we only describe the relatively straightforward
extension we used to model multiple strains per functional group.

Each of the strains within each functional groups has its own parameter
set and state variable. Thus a system with nine strains within each
functional group has 27 state variables, plus another four (one for each
of the four chemical substrates). Other than having potentially
different parameter values (e.g.~maximum growth rate and tolerance
parameters) the dynamics of the strains in a functional were described
by the same equations. Hence, when parameterised with multiple strains
per functional group, but no variation among strains (within functional
groups), the system behaves identically to when there is only one strain
per functional group (see the \texttt{microxanox} R package \emph{User
Guide} for a demonstration).

The only other necessary modification was to include summation terms in
each of the four ordinary differential equations describing the rates of
change of the four chemical substrates. This was required in order to
appropriately account for the production and consumption of these
substrates by multiple strains.

@@

In this section we describe how we created variability among the
multiple strains within functional groups. Since the process was the
same for each functional group, we usually do not in this description
subscript parameter names by functional group (though there were
functional group specific parameters in all simulations). The
descriptions in this section gradually become more mathematical; the
next paragraph alone should provide the information required to
understand all other elements of this report.

We introduced among-strain variability in maximum growth rate (parameter
\(G_{max}\)) and in the tolerance to the relevant inhibiting chemical
substrate (parameter \(h\)). Among-strain variability was created such
that there was a tradeoff between maximum growth rate and tolerance
(i.e.~high tolerance coinciding with low maximum growth rate)
(e.g.~Figure @ref(fig:tradeoff). Our implementation of the variation and
tradeoff contain the assumption of a linear tradeoff of log-log
transformed among-strain variation.

@@

{\centering \includegraphics[width=0.5\linewidth]{article_files/figure-latex/tradeoff-1}

@@

\caption{An example of among-strain variation in growth rate and tolerance, and the tradeoff in the variation.}\label{fig:tradeoff}
\end{figure}

For further understanding, please consider a reference maximum growth
rate parameter \(G_{max}\) and reference tolerance parameter \(h\).
Furthermore, consider the meta-parameters \(Div_{G_{max}}\) and
\(Div_{h}\), which determine the amount of variation among \(n\)-strains
in the growth rate and tolerance parameters. To create a tradeoff, the
two meta-parameters were always of different sign.

The among strain variation was calculated such that the growth rate of
strain \(i = 1\) was a factor \(1/(2^{Div_{G_{max}}})\) of the reference
growth rate, and the growth rate of the \(i = n\) strain was a factor
\(2^{Div_{G_{max}}}\) of the reference growth rate. The growth rate of
the \(i = \{2,...n-1\}\) other strains was the reference growth rate
multiplied by a factor that was equally distributed between the factor
of the \(i = 1\) and \(i = n\) strain.

Hence, for example, with the reference growth rate \(G_{max} =\) 0.05,
among strain variation of \(Div_{G_{max}} =\) 1, and nine strains
(\(n =\) 9), then the growth rates of the nine strains are 0.025, 0.03,
0.035, 0.042, 0.05, 0.059, 0.071, 0.084, 0.1 (Figure @ref(fig:tradeoff).

(Put another way, the growth rate parameter of the i-th of the \(n\)
strains in a functional group was calculated as
\(G_{max, i} = D_i * G_{max}\) where D\_i is the i-th element of the set
\(D = \{2^{f(x)Div_{G_{max}}} | x \in \mathbb{N}, x \leq n\}\), where
\(f(x) = (x-(n+1)/2) / (( n-1)/2)\).)

@@

\caption{No diversity}\label{fig:ss_novar}
\end{figure}

@@

\caption{Diversity in all functional groups}\label{fig:ss_var}
\end{figure}

@@

\caption{Diversity in none (solid) and all functional groups.}\label{fig:ss_var2}
\end{figure}

@@

\caption{Diversity in CB functional group.}\label{fig:CB_var1}
\end{figure}

@@

\caption{Diversity in none (solid) and CB functional groups.}\label{fig:CB_var2}
\end{figure}

\hypertarget{figure-sb-stable-states-with-variation-and-tradeoff}{%
\subsection{Figure: SB Stable states with variation and
tradeoff}\label{figure-sb-stable-states-with-variation-and-tradeoff}}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/SB_var1-1}

\caption{Diversity in SB functional group.}\label{fig:SB_var1}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/SB_var2-1}

\caption{Diversity in none (solid) and SB functional group.}\label{fig:SB_var2}
\end{figure}

\hypertarget{figure-pb-stable-states-with-variation-and-tradeoff}{%
\subsection{Figure: PB Stable states with variation and
tradeoff}\label{figure-pb-stable-states-with-variation-and-tradeoff}}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/PB_var1-1}

\caption{Diversity in PB functional group.}\label{fig:PB_var1}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/PB_var2-1}

\caption{Diversity in none (solid) and PB functional group.}\label{fig:PB_var2}
\end{figure}

@@

\caption{Diversity in SB and PB functional groups.}\label{fig:SBPB_var1}
\end{figure}

@@

\caption{Diversity in none (solid) and SB and PB functional groups.}\label{fig:SBPB_var2}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ss_var3-1}

@@

\caption{your caption}\label{fig:ss_var3}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{article_files/figure-latex/ss_var4-1}

@@

\caption{your caption}\label{fig:ss_var4}
\end{figure}

\hypertarget{figure-temporal-switches}{%
\subsection{Figure: Temporal switches}\label{figure-temporal-switches}}

Maybe\ldots{} to be added. Need to go both ways (oxic to anoxic, and
anoxic to oxic).

@@

>>>>

This report concerns how biodiversity influences the strength and nature
of ecosystem responses to environmental change (figure 1; panel(a) three
ovals, environment oval with arrow to ecosystem property oval, with
biodiversity oval pointing the arrow between environment and ecosystem
property; panel (b) shows linear response with different gradients,
large being with low diversity; panel (c) shows a linear and a
non-linear response starting and ending at the same place; panel (d)
shows a hysteretic response, with greater or less range of hysteresis).
Strong responses of ecosystems to environmental change imply high
sensitivity and lack of resistance to environmental change, and higher
temporal variability in the face of varying environmental conditions.
Greater non-linearity in ecosystem responses increases the potential for
surprises, such as when an ecosystem that was resistant to an
environmental change becomes unexpectedly sensitive to further change.
The presence and extent of hysteresis in ecosystem responses to
environmental driver gradients concerns how readily are ecosystem
responses reversed when environmental change is reversed.

@@

\begin{itemize}
\item
  Greater functional response diversity may decrease the likelihood of a
  catastrophic ecosystem response (Figure 2a in Dakos et al. (2019)).
  Here, we interpret this prediction to be that greater functional
  response diversity decreases the non-linearity of ecosystem response
  to environmental driver change, including the extreme non-linearity
  that can occur when an ecosystem ``flips'' from one state to another
  one.
\item
  Greater functional trait diversity can make a tipping point occur only
  with greater change in an environmental driver (Figure 2a in Dakos et
  al. (2019)). I.e. functional trait diversity can increase
  resilience--the extent of environmental change required to push the
  ecosystem into another state (Holling (1973)). This could occur if
  greater functional diversity equates to the presence of organisms with
  greater tolerance to the environmental driver Dakos et al. (2019).
\item
  On the other hand, the opposite could occur: greater functional
  diversity could make a tipping point occur with less change in an
  environmental driver (i.e.~may cause lower resilience) (Figure 2c in
  Dakos et al. (2019)). This could occur if there is selection for
  greater tolerance that causes, incidentally, selection for lower
  growth rate, perhaps due to a tradeoff between the two traits.
\item
  Finally, the amount of functional diversity may affect the ``recovery
  trajectory,'' i.e.~the relationship between ecosystem state and the
  environmental driver when the driver change is reversed Dakos et al.
  (2019). Put another way, functional diversity may increase or decrease
  the amount of hysteresis, and may even result in a recovery with lower
  non-linearity / no tipping-point (Figure 3 in Dakos et al. (2019)).
\end{itemize}

These explanations/predictions were based on the verbal and graphical
argumentation Dakos et al. (2019). There were no mathematical models of
ecosystem responses to environmental change behind the
explanations/predictions. Ceulemans, Wojcik, and Gaedke (2021), in their
article \emph{Functional diversity alters the effects of a pulse
perturbation on the dynamics of tritrophic food webs}, did employ a such
a mathematical model {[}text to be developed{]}.

@@

\begin{itemize}
\tightlist
\item
  To provide an answer to the question of how trait variation affects
  the resilience and hysteresis of ecosystem response to an
  environmental change.
\end{itemize}

@@

\begin{itemize}
\item
  Mathematical modelling of a microbial ecosystem.
\item
  Trait diversity manipulation (can be considered selection among
  strains, or selection among species). (Do not include mutation or
  recombination).
\item
  Variation in amount of among strain variation in two traits, no
  variation in number of strains, no variation in strength or shape of
  the of the tradeoff.
\item
  Illustration of the dynamical response to gradual environmental
  change.
\item
  Measurement of the effect trait diversity on how the steady state(s)
  of the system varies with environmental conditions.
\end{itemize}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/owenpetchey/Desktop/microxanox/diversity_envresp1/manuscript/figures/ecosystem_network}

@@

Bush et al (2017) contains an accessible and complete description of the
model of the ecosystem, including rate equations, parameter values, and
initial conditions. These features are also represented and documented
in the \texttt{microxanox} R package (\textbf{cite the package}).
Therefore, here we only describe the extension we used to model multiple
strains per functional group.

Each of the strains within each functional groups has its own parameter
set and state variable. Thus a system with nine strains within each
functional group has 27 state variables, plus another four (one for each
of the four chemical substrates). Other than having potentially
different parameter values (e.g.~maximum growth rate and tolerance
parameters) the dynamics of the strains in a functional were described
by the same equations. Hence, when parameterised with multiple strains
per functional group, but no variation among strains (within functional
groups), the system behaves identically to when there is only one strain
per functional group (see the \texttt{microxanox} R package \emph{User
Guide} for a demonstration). The only other necessary modification was
to include summation terms in each of the four ordinary differential
equations describing the rates of change of the four chemical
substrates. This was required in order to appropriately account for the
production and consumption of these substrates by multiple strains.

@@

@@

{\centering \includegraphics[width=1\linewidth]{/Users/owenpetchey/Desktop/microxanox/diversity_envresp1/manuscript/figures/diversity_example}

@@

\caption{your caption}\label{fig:diversity_examples}
\end{figure}

In this section we describe how we created variability among the
multiple strains within functional groups. In brief, in each functional
group we created multiple distinct strains that differed in their
maximum growth rate and tolerance to an inhibiting substrate, with a
trade-off between the two (e.g.~higher tolerance assocated with lower
maximum growth rate). An example of the resulting variation is displayed
in Figure \ref(fig:diversity\_example)).

A more detailed account of the creation of strains follows now.
Introducing variation among multiple strains within a functional group
required a number of decisions:

\begin{itemize}
\tightlist
\item
  \textbf{The number of strains.} The symbol for this is \(n\)
  subscripted with the functional group code (CB = cyanobacteria, SP =
  sulphate reducing bacteria, and PB = phototrophic sulphur bacteria).
  E.g. \(n_{CB}\) is the number of strains of cyanobacteria. {[}add why
  we only ever had 9{]}.
\item
  \textbf{Which traits to introduce variation in.} We chose the maximum
  growth rate and the tolerance of growth to an inhibiting substrate.
  E.g. maximum growth rate of the cyanobacteria \(G_{max,CB}\), and the
  tolerance of cyanobacterial growth to reduced sulphur \(h_{SR,CB}\).
\item
  \textbf{The amount of among strain variation.} The symbol for this is
  \(Div\) (from \emph{Div}ersity) subscripted by the trait/parameter and
  the functional group. E.g. * \(Div_{G_{max,CB}}\) is the amount of
  among strain variation in the maximum growth rate \(G_{max}\) of the
  cyanobacteria \(CB\).
\item
  \textbf{The correlation between variation in different traits.} This
  was always negative, so there was a tradeoff. Our implementation of
  the variation and tradeoff contain the assumption of a linear tradeoff
  of log-log transformed among-strain variation.
\item
  \textbf{The relative amount of among strain variation in different
  functional groups.} E.g. were strains of cyanobacteria very variable
  compared to strains of the other functional groups. We performed
  several numerical experiments, varying which functional groups had
  more or less among strain variation.
\item
  \textbf{The distribution of trait variation.} We chose to have among
  strain variation distributed evenly on a log scale.
\end{itemize}

The parameters therefore necessary were:

\begin{itemize}
\tightlist
\item
  \(n_{CB}\) - number of strains of cyanobacteria.
\item
  \(n_{SB}\) - number of strains of sulphate reducing bacteria.
\item
  \(n_{PB}\) - number of strains of phototrophic sulphur bacteria.
\item
  \(Div_{G_{max,CB}}\) - amount of among strain variation in the maximum
  growth rate \(G_{max}\) of the cyanobacteria \(CB\).
\item
  \(Div_{h_{SR,CB}}\) - amount of among strain variation in the
  tolerance \(h\) of cyanobacteria \(CB\) to reduced sulpur compounds
  \(SR\).
\item
  \(Div_{G_{max,SB}}\) - amount of among strain variation in the maximum
  growth rate \(G_{max}\) of the sulphate reducing bacteria \(SB\).
\item
  \(Div_{h_{O,SB}}\) - amount of among strain variation in the tolerance
  \(h\) of sulphate reducing bacteria \(S\) to oxygen \(O\).
\item
  \(Div_{G_{max,PB}}\) - amount of among strain variation in the maximum
  growth rate \(G_{max}\) of the phototrophic sulphur bacteria \(PB\).
\item
  \(Div_{h_{O,PB}}\) - amount of among strain variation in the tolerance
  \(h\) of phototrophic sulphur bacteria \(PB\) to oxygen \(O\).
\end{itemize}

For example, cyanobacteria, and a reference maximum growth rate
parameter \(G_{max,CB}\) and reference tolerance parameter
\(h_{SR,CB}\). Recall that amount of among strain variation is
controlled by the two meta-parameters \(Div_{G_{max,CB}}\) and
\(Div_{h_{SR,CB}}\). To create a tradeoff, the two meta-parameters were
always of different sign. The among strain variation was calculated such
that the growth rate of strain \(i = 1\) was a factor
\(1/(2^{Div_{G_{max,CB}}})\) of the reference growth rate, and the
growth rate of the \(i = n\) strain was a factor
\(2^{Div_{G_{max,CB}}}\) of the reference growth rate. The growth rate
of the \(i = \{2,...n-1\}\) other strains was the reference growth rate
multiplied by a factor that was equally distributed between the factor
of the \(i = 1\) and \(i = n\) strain. This implementation of the
variation and trade-off contains the assumption of a linear trade-off of
log-log transformed among-strain variation.

Hence, for example, with the reference growth rate \(G_{max,CB} =\)
0.05, among strain variation of \(Div_{G_{max,CB}} =\) 1, and nine
strains (\(n =\) 9), then the growth rates of the nine strains are
0.025, 0.03, 0.035, 0.042, 0.05, 0.059, 0.071, 0.084, 0.1 (Figure to be
added).

(Put another way, the growth rate parameter of the i-th of the \(n\)
strains in a functional group was calculated as
\(G_{max, i} = D_i * G_{max}\) where \(D_i\) is the i-th element of the
set \(D = \{2^{f(x)Div_{G_{max}}} | x \in \mathbb{N}, x \leq n\}\),
where \(f(x) = (x-(n+1)/2) / (( n-1)/2)\).)

@@

\caption{Steady states of the ecosystem with no trait diversity, i.e. identical trait values for the strains within a functional group.}\label{fig:ss_novar}
\end{figure}

@@

\caption{Steady states of the ecosystem with diversity in all functional groups}\label{fig:ss_var}
\end{figure}

@@

\caption{Steady states of the ecosystem without any diversity (solid line) and with diversity in all three functional groups.}\label{fig:ss_var2}
\end{figure}

@@

\caption{Steady states of the ecosystem with diversity in only the cyanobacteria functional group.}\label{fig:CB_var1}
\end{figure}

@@

\caption{Steady states of the ecosystem without any diversity (solid line) and with diversity in only the cyanobacteria functional group.}\label{fig:CB_var2}
\end{figure}

@@

\caption{Steady states of the ecosystem with diversity in only the two sulphur bacteria functional groups.}\label{fig:SBPB_var1}
\end{figure}

@@

\caption{Steady states of the ecosystem without any diversity (solid line) and with diversity in only the sulphur bacteria functional group.}\label{fig:SBPB_var2}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/owenpetchey/Desktop/microxanox/diversity_envresp1/manuscript/figures/extent_of_bistab1}

@@

\caption{Effect of increasing amounts of trait diversity on the extent of the environmental gradient for which bistability (i.e. alternative stable states) is present, when different combinations of the three functional groups contain any diversity.}\label{fig:stab1}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{/Users/owenpetchey/Desktop/microxanox/diversity_envresp1/manuscript/figures/extent_of_bistab2}

@@

\caption{Effect of increasing amounts of trait diversity on the upper and lower bound of the environmental gradient for which bistability (i.e. alternative stable states) is present, when different combinations of the three functional groups contain any diversity.}\label{fig:stab2}
\end{figure}

@@

We observed that greater trait diversity among the organisms dominating
the prevailing ecosystem state can lead to greater resilience of that
state. We also observed that greater trait diversity among the organisms
that occur in an different ecosystem state can reduce resilience.

The effect of diversity occurs because with greater trait diversity
there is a organism with greater tolerance to the deteriorating (from
the perspective of the dominant organisms). This greater tolerance means
that a greater amount of environmental deterioration is required to
cause the ecosystem to flip into the alternative state.

In the case study, with three functional groups, the presence or absence
of diversity in the different functional groups influences
resilience\ldots{} with only diversity in the CB\ldots{}
